\documentclass[12pt]{article}
\input{typesetting.tex}
\author{Yingbo Ma\\ \tt{yingbom@uci.edu}}
\date{March 2018}
\title{Stiffness Detection and Automatic Switching Algorithms for
OrdinaryDiffEq.jl and Tooling for ModelingToolkit.jl \\
\large{Proposal for the GSoC 2018 Project}}

\begin{document}
\maketitle
\tableofcontents

\section{Motivation \& The Project}
There are two projects that I propose for the Google Summer of Code (GSoC)
Program, one is \textit{stiffness detection and automatic switching
algorithms}, and the other is \textit{tooling of} \texttt{ModelingToolkit.jl}. By
the end of the program, I hope to have implemented a production ready automatic
stiffness detection algorithms, intermediate representation (IR) manipulating
tools, and index reduction algorithm for differential algebraic equation (DAE).

\subsection{Goal 1: Implement stiffness detection and automatic switching
algorithms} \label{goal1}
An ordinary differential equation (ODE) is an equality relationship among a
function $y(t)$, its independent variable $t$ and its derivatives. An $n$th
order ODE can be written as
\begin{equation}
  F(t, y(t), y'(t), \cdots, y^{(n-1)}(t)) = 0.
\end{equation}
ODEs are very important, because they are raised from every branch of science.
Most of the ODEs which people encounter have no analytical solution. Hence,
people must solve them numerically. Stiff ODEs are a kind of problem which is
very hard to solve by most explicit numerical methods, however, they can be
solved by implicit methods like the family of Rosenbrock methods or backward
differentiation formulae (BDF) without struggle. The problem is that the
methods which are good for stiff ODEs are not fast, as they have to solve for
non-linear or linear systems in each time step. Yet, most stiff problems are
not stiff throughout the whole domain of interest. This leads to the motivation
of my first project, \textit{stiffness detection and automatic switching
algorithms}. Conventionally, one uses only a single algorithm to solve for the
entire domain of interest. Thus, one is forced to use a stiff algorithm,
although there is only a very small interval in the domain that is stiff. The
auto-switching algorithm can solve this dilemma and raise the efficiency of
solving ODEs, because stiff algorithms will only be applied to the stiff part
of the domain, and non-stiff algorithms will solve the rest of the problem
quickly. This project is beneficial for most people who use the
\textit{JuliaDiffEq} ecosystem, because it removes the burden of having to test
multiple algorithms for one problem and chose the more efficient one. It can
also help to save much computation time for huge ODE systems, e.g. time
stepping of a PDE discretization.

\subsubsection{Implementation Details}
Higham and Trefethen~\cite{stiffode} pointed out that an ODE system is stiff
when the pseudospectrum of its $J$ has big radii while it has a small abscissa,
where $J$ denotes a Jacobian matrix. Computing the pseudospectra of a linear
operator is very costly, so in practice, it makes sense to only calculate the norm
or the approximated largest eigenvalue of $J$.

The simplest automatic stiffness detection on explicit Runge-Kutta algorithms
is proposed by Ernst Hairer~\cite{hairer2}. One can use
\begin{equation}
  \abs{\lambda} = \frac{\norm{f(y+v, t)-f(y, t)}}{\norm{v}}
\end{equation}
to approximate the dominant eigenvalue of the Jacobian matrix of the function
$f(y, t)$ when $\norm{v}$ is small. For explicit Runge-Kutta methods like
Dormand-Prince and Tsitouras $5/4$ pairs, there is the relation $c_6 = c_7 =
1$. Hence, one can use
\begin{equation}
  \varrho = \frac{\norm{k_7-k_6}}{\norm{g_7-k_6}}\qqtext{where } k_i = f(g_i,
  t_0 + c_i\cdot \Delta t)
\end{equation}
to produce a decent approximation of the most significant eigenvalue.
Naturally, one can classify a problem to be stiff when $\Delta
t\cdot\varrho/S \ll 1$. Where $S$ is the width of the stability domain of the
Runge-Kutta algorithm. Currently, the implementation reads

\begin{lstlisting}
g6 = uprev+dt*(a61*k1+a62*k2+a63*k3+a64*k4+a65*k5)
k6 = f(g6, p, t+dt)
g7 = u
rho = integrator.opts.internalnorm(k7 - k6)/
      integrator.opts.internalnorm(g7 - g6)
stiffness = rho*dt/3.3
\end{lstlisting}

and here is a demonstration of this automatic stiffness detection method with
the infamous Dormand-Prince $5/4$ integrator. The ODEs that the integrator
solves are
\begin{align}
  y'_1 &= y_2\\
  y'_2 &= ((1-y_1^2)y_2-y_1) / \varepsilon
\end{align}
where $\varepsilon$ is $0.003$.

\input{img/vdp.tex}

\subsection{Goal 2: Tooling for \texttt{ModelingToolkit.jl}}
ModelingToolkit.jl~\cite{modelingtoolkit} is an intermediate representation
(IR) of computational graphs for scientific computing problems. My second goal
during this GSoC project is to implement expression parsing tools and utilities
for IR manipulation in the repository. For instance, user
Borisov~\cite{parsing} expression parsing of the form
\begin{lstlisting}
ex = quote
    v1 = k1*A*comp1
    dA = -v1/comp1
    dB = v1/comp1
end
params = (:k1, :comp1)
depend_vars = (:A, :B)
\end{lstlisting}
which requires tools for expression parsing.

\subsection{Goal 3: Implement DAE index reduction algorithms}
My third goal is to implement index reduction algorithms for differential
algebraic equations (DAEs). DAEs are raised naturally from Euler-Lagrange
equations with constraints and many modeling problems, and its general form is
\begin{equation}
  F(t, \vec{x}, \frac{d\vec{x}}{dt}) = \vec{0}.
\end{equation}
However, numerical DAE solvers can only handle Hessenberg index-1 and
Hessenberg index-2 forms,
\begin{align}
  &\textbf{Hessenberg index-1}
  \begin{dcases}
    y_i' &= f_i(t,y_i,z_i)\\
    0 &= g_i(t,y,z)
  \end{dcases}\\
  &\textbf{Hessenberg index-2}
  \begin{dcases}
    y'_i &= f_i(t,y_i,z_i)\\
    0 &= g_i(t,y)
  \end{dcases}
\end{align}
so one has to reduce the index of the DAE to index-1 or index-2 form. However,
it is difficult and cumbersome to reduce the indices of a DAE system by hand.
The \texttt{ModelingToolkit.jl} package has enough information about a DAE system,
because it has its own computational graph of the system. This project benefits
people who work on modelings and dynamical systems when the solutions' local
coordinates are complicated and not trivial to convert to.

\subsection{Stretch Goals}
\begin{itemize}
  \item Implement utilities for plotting Poincar\'e sections.
  \item Implement \texttt{IIF1} and \texttt{IIF2} methods for solving
    semi-linear partial differential equations (PDEs)
\end{itemize}

\subsection{Potential Hurdles}
The potential hurdles that I expect are mostly about design decisions. I
believe that I can overcome this hurdle by learning from my potential
mentors'--- Jiahao and Chris'---suggestions. For instance, when I was
implementing the order reduction algorithm for systems of ODEs, Chris taught me
that this kind of algorithms should be implemented as a transformation that is
\[
  f: \texttt{DiffEqSystem}\mapsto \texttt{DiffEqSystem},
\]
so that it fits into the design of the \texttt{ModelingToolkit.jl} package, and
makes it more flexible and extendible~\cite{i49}.

I also might will have struggle to use \texttt{Git}. I have messed up a pull
request by a bad \texttt{git rebase}. However, I am confident that I can subdue
this problem by being extra careful when I merge or rebase in a \texttt{Git}
repository.

\subsection{Potential Mentors}
Jiahao Chen (\texttt{@jiahao}) is going to be my primary mentor, and
Christopher Rackauckas (\texttt{@ChrisRackauckas}) is going to my secondary
mentor.

\section{Milestones}
There are many different aspects of a coding project like features, API
usability and documentations. I will prioritize the robustness aspect over
everything else in the beginning of the GSoC project, since the correctness is
very critical for numerical analysis softwares. I will achieve robustness by
adding extensive tests and benchmarks for all of my pull requests. I will shift
my attention to API usability after I finish all the features, as a software
must be friendly to the users to gain popularity. After I stabilize the API, I
will move my focus to documentations. I will not only write the standard
documentations in \url{http://docs.juliadiffeq.org/latest/}, but also more
tutorials in the \texttt{DiffEqTutorials.jl} repository. Here is my tentative
plan for the GSoC project.

\paragraph{Community Bonding: April 5 --- May 14}
I am going to read more literature about the stiffness of ODEs, and discuss
many possible implementations for my mentors.

\paragraph{First Milestone --- Stiffness Detection Algorithms: May 14 --- June
11}
I should have implemented most of the stiffness detection algorithms by June
11, and most of the methods switching firework should also be finished as well.
At this stage, I will be ready for dive into benchmarks, and try to find the
best methods switching strategy, e.g. switch to one of the Rosenbrock methods
after hitting 15 successive stiff test positives.

\paragraph{First Evaluation: June 11 --- June 15}
At this point, I should have completed my first goal~\ref{goal1}, and I should
clean up the code that I have produced during the last period.

\paragraph{Second Milestone --- \texttt{ModelingToolkit.jl} IR Tooling: June 15
--- July 9}
I will turn my focus to the \texttt{ModelingToolkit.jl} repository at this
stage, but I will still try to optimize the methods switching strategy that is
completed in the previous milestone. I will also fix some bugs in the strategy
and the algorithms, since after stiffness detection and automatic switching
algorithms are tagged in the \texttt{OrdinaryDiffEq.jl} repository, users will
perhaps file bug reports about those algorithms. I should have finished
expression parsing by July 9.

\paragraph{Second Evaluation: July 9 --- July 13}
I am going to clean up my code in the \texttt{ModelingToolkit.jl} repository.
After filling the second evaluation form, I should have finished all my GSoC
goals at this step.

\paragraph{Third Milestone --- DAE index reduction algorithms \& Stretch Goals:
July 13 -- August 6}
I am going to work on DAE index reduction algorithms and stretch goals at this
stage. DAE index reduction algorithms are not very complicated and I think I
should have time to work on my stretch goals if everything goes well.

\paragraph{Finial Evaluation: August 6 --- August 14}
I will complete my finial evaluation report and clean up my code in this final
stage of the GSoC project.

\section{Code Portfolio}
Here is this a list of contributions that I did during the previous GSoC
project. \url{https://gist.github.com/YingboMa/65dbe1a43564ab36740858846
7ccb671}

\section{Deliverables}
\begin{itemize}
  \item Stiffness Detection and Automatic Switching Algorithms in
    the \texttt{OrdinaryDiffEq.jl} repository.
  \item Tools for manipulating IR in the \texttt{ModelingToolkit.jl}
    repository.
  \item DAE index reduction algorithms in the \texttt{ModelingToolkit.jl}
    repository.
\end{itemize}

\section{About Me}
I am a physics major freshman in the University of California, Irvine. I am a
former GSoC student who worked on boundary value problem solvers,
Runge-Kutta-Nystr{\"o}m methods and symplectic integrators.

\subsection{Academic Details}
\begin{itemize}
  \item University: University of California, Irvine
  \item Major: Physics (Freshman)
  \item GPA: 4.0
\end{itemize}

\subsection{Contact Information}
\paragraph{Email:} mayingbo5@gmail.com
\paragraph{GitHub:} YingboMa

\section{Summer Logistics}
I expect to be able to work over 30 hours per week throughout the summer. I do
not have any summer courses nor other jobs, so I can put most of my attention
toward this GSoC project. Overall, I will be able to put 350-400 hours into
this project. If possible I plan to visit MIT Julia Lab during this GSoC
project.

\bibliography{cite}
\end{document}
